---
title: "Pre-fire vegetation"
author: "Nils Rietze"
date: "2025-02-03"
output:
  html_document:
    code_folding: "hide"
    theme: sandstone
    highlight: tango
---
```{r setup, include=FALSE}
library(terra)
library(tidyterra)
library(spatstat.utils)
library(gt)
library(tidyverse)
library(cowplot)
library(patchwork)
set.seed(10)

knitr::opts_chunk$set(
    # This should allow Rmarkdown to locate the data
    root.dir = rprojroot::find_rstudio_root_file()
)
```

# User functions
```{r}
read_hls_data_frames <- function(index_name, UTM_TILE_ID,year,
                                 sample_points, dnbr_sample){
  filename <- sprintf("%s_sampled_%s_%s.csv",index_name,UTM_TILE_ID,year)
  
  df <- read.csv2(paste0(HLS_DIR,filename)) %>% 
    select(-1) %>% 
    mutate(dNBR = dnbr_sample$dNBR) %>% 
    as_tibble() %>% 
    mutate(burn_date = sample_points$burn_date)
  
  if (index_name == "LST"){
    fmt = "X%Y.%m.%d"
  } else{
    fmt = "X%Y.%m.%d.%H.%M.%S"
  }
  
  df_long <- df %>%
    mutate(ObservationID = 1:nrow(df)) %>%  # Add a column for observation IDs (row numbers)
    gather(key = "Time", value = index_name, 
           -c(ObservationID,dNBR,burn_date)) %>% 
    mutate(Time = as.POSIXct(Time, format = fmt)) 
  
  return(df_long)
}
```


# 0. Load and prepare data
```{r, include=FALSE}
# Config, loading and preparing data
HLS_DIR <- "~/data/raster/hls/"
index_name = "NDMI"
TEST_ID <- 14211 # fire ID for part of the large fire scar

UTM_TILE_ID <- "54WXE"
year <- 2020
severity_index <- "dNBR"

pct_cutoff <- 0.5

# Load dNBR raster
dnbr <- rast(
  sprintf("~/data/raster/hls/severity_rasters/%s_%s_%s.tif",
          UTM_TILE_ID, year,severity_index)) * 1000

# Load features (fire perimeters and ROIs)
fire_perimeters <- vect(
  "~/data/feature_layers/fire_atlas/viirs_perimeters_in_cavm_e113.gpkg"
)

selected_fire_perimeter <- fire_perimeters %>% 
  filter(fireid  == TEST_ID) %>%
  project(crs(dnbr))

# Buffer the selected fire perimeter
fire_perimeter_buffered <- buffer(selected_fire_perimeter, 1200)

dnbr_in_perimeter <- dnbr %>% 
  mask(fire_perimeter_buffered, updatevalue = NA) %>% 
  crop(fire_perimeter_buffered)

# Load sampled points (with associated burn dates)
sample_points <- vect("~/data/feature_layers/sample_points_burn_date.gpkg") %>% 
  project(crs(dnbr)) %>% 
  mutate(ObservationID = 1:nrow(.))

# Extract data at random points 
dnbr_sample <- terra::extract(dnbr, sample_points,
                              ID = FALSE, xy = TRUE)

# Load NDMI raster for nr. of observations
df_ndmi <- read_hls_data_frames("NDMI", UTM_TILE_ID,year,sample_points, dnbr_sample) %>% 
  rename("NDMI" = index_name)

# Load filtered and formatted dataframe
fn_filtered_df <- sprintf("data/tables/merged_filtered_%sth_pctile.csv",pct_cutoff*100)

df_filtered <- read.csv2(fn_filtered_df) %>% 
    mutate(date = as.Date(date),
           burn_date = as.Date(burn_date))
```

# 1. Descriptive statistics
## a. Distribution of dNBR
How does the dNBR distribution look for the entire fire event perimeter? Below is the histogram and the corresponding dNBR raster:
```{r}
binwidth <- 20

# Plot dNBR map
p1 <- ggplot() +
  geom_spatraster(data = dnbr_in_perimeter) +
  scale_fill_viridis_c(option = "inferno",
                       na.value = "white",
                       name = "dNBR") +
  theme_cowplot()

# Plot dNBR for this map
p2 <- ggplot(data = dnbr_in_perimeter, aes(x = dNBR)) +
  geom_histogram(binwidth = binwidth, color = "black", fill = "indianred1") +
  xlim(c(-500,1000)) +
  labs(
    title = "Histogram of dNBR",
    x = "dNBR Values",
    y = "Frequency"
  ) +
  theme_cowplot()

(pg <- p1 + p2)

# ggsave2(pg,filename = "figures/expl_dNBR_hist.png",width = 8,height = 6,bg = "white")
```

and what does it look like for the sampled points? Remember the points were sampled using a stratification from a binned dNBR histogram, so the histograms should be similar.
```{r}
# Plot distribution of total raster vs. sample points
p3 <- ggplot(data = dnbr_sample, aes(x = dNBR)) +
  geom_histogram(binwidth = binwidth, color = "black", fill = "orange") +
  xlim(c(-500,1000)) +
  labs(
    title = "Histogram of dNBR (sampled)",
    x = "dNBR Values",
    y = "Frequency"
  ) +
  theme_cowplot()

(pg <- p2 / p3 )
# ggsave2(pg, filename = "figures/expl_dNBR_hist_vs_sampled.png",
#         width = 8, height = 6,bg = "white")
```

## b. Check nr. of available observations per point
Let's see how many non-NA observations each sampled point has:
```{r}
# get nr. of valid observations per point
valid_counts_by_id <- df_ndmi %>%
  group_by(ObservationID) %>%
  summarise(valid_count = sum(!is.na(NDMI)))

# Add this info to the spatvect point layer
sample_points_filtered <- sample_points %>% 
  merge(valid_counts_by_id, by = "ObservationID")

# Plot map and show number of observations per points
(p <- ggplot() +
    # geom_spatraster(data = dnbr_in_perimeter, aes(fill = dNBR)) +
    geom_spatvector(data = sample_points_filtered, aes(color = valid_count)) +
    scale_color_viridis_c(option = "magma",
                         na.value = "transparent") +
    labs(fill = "Number of non-masked observations") +
    theme_cowplot()
)

# ggsave2(p, 
#         filename = "figures/Nobs_randompoints_map.png",
#         bg = "white",width = 10, height = 8)
```

Here are the summary statistics for the nr. of valid observations
```{r}
# Report summary stats for pre-fire observations
df_filtered %>%
  group_by(ObservationID) %>%
  filter(BeforeBurnDate) %>%
  summarise(
    n_observations = n()
  ) %>%
  summarise(
    mean_obs = mean(n_observations),
    median_obs = median(n_observations),
    min_obs = min(n_observations),
    max_obs = max(n_observations)
  )
```

What does the histogram of nr. of non-na observations look like? I added a vertical line at a percentile-threshold.
```{r, include=FALSE}

# Create dataframe with nr. of valid observations per point
df_nobs <- df_ndmi %>% 
  select(-dNBR) %>% 
  group_by(ObservationID) %>% 
  summarize(valid_count = sum(!is.na(NDMI)))

# Define percentile of nr. of observations to use as cutoff
pct_cutoff <- 0.75
thr_nobs <- quantile(df_nobs$valid_count,pct_cutoff , na.rm = TRUE)

(ggplot(df_nobs) +
  geom_histogram(aes(x = valid_count),binwidth = 5) +
  geom_vline(xintercept = thr_nobs,
             color = "red", linetype = "dashed", linewidth = 1) + 
  geom_text(x = thr_nobs-10, y = 300, 
            label = sprintf("%sth percentile = %s",
                            pct_cutoff*100, round(thr_nobs, 2)),
            color = "red", vjust = -1) +
  labs(title = "Number of non-NA observations",
       subtitle = sprintf("Threshold of valid data (%sth percentile)",
                          pct_cutoff*100)) +
  theme_cowplot())

# ggsave2(sprintf("figures/Histogram_%s_observations_%sth_pctile.png",
#                 index_name,pct_cutoff*100),
#         bg = "white",width = 10, height = 8)

```

## b. Points filtered by nr. of non-na observations
Where are these filtered points? Here's a map of them:
```{r}
# Plot where these filtered points sit
sample_points_filtered <- sample_points %>% 
  filter(ObservationID %in% df_filtered$ObservationID)

(p <- ggplot() +
  geom_spatraster(data = dnbr_in_perimeter, aes(fill = dNBR)) +
  geom_spatvector(data = sample_points_filtered) +
  scale_fill_viridis_c(option = "magma",
                       na.value = "transparent") +
  labs(fill = "dNBR") +
  theme_cowplot()
)

# ggsave2(p, 
#         filename = sprintf("figures/Location_randompoints_%sth_pctile.png",pct_cutoff*100),
#         bg = "white",width = 10, height = 8)

```

Is the dNBR distribution for this final subset still similar to the original dNBR distribution?
```{r}
p1 <- ggplot(data = dnbr_in_perimeter, aes(x = dNBR)) +
  geom_histogram(binwidth = binwidth, color = "black", fill = "indianred1") +
  xlim(c(-500,1000)) +
  labs(
    title = "Histogram of dNBR in fire perimeter",
    x = "dNBR Values",
    y = "Frequency"
  ) +
  theme_cowplot()

p2 <- df_filtered %>% 
  group_by(ObservationID) %>% 
  summarise(dNBR = first(dNBR)) %>%
  ggplot(data = ., aes(x = dNBR)) +
  geom_histogram(binwidth = binwidth, color = "black", fill = "indianred1") +
  xlim(c(-500,1000)) +
  labs(
    title = "Histogram of dNBR in subset",
    x = "dNBR Values",
    y = "Frequency"
  ) +
  theme_cowplot()

(pg <- p1 + p2)
```
The sampled subset cutting off points with fewer than 80 observations seems to resemble the overall dNBR distribution but cuts off high-severity values. So probably, the cutoff is too conservative.

The question now is a trade-off of capturing the distribution better or avoiding more spatial autocorrelation.

## c. Check spatial autocorrelation of final point sample
```{r}
library(spatstat)
library(spdep)

knn <- knn2nb(knearneigh(dnbr_sample[c("x","y")], k = 5))

# Convert neighbors to weights
weights <- nb2listw(knn, style = "W")

# Compute Moran's I
moran_i <- moran.test(dnbr_sample$dNBR, listw = weights)

(ggplot() +
  geom_spatraster(data = mask(dnbr,fire_perimeter_buffered)) +
  scale_fill_viridis_c(option = "inferno",
                       na.value = "white",
                       name = "dNBR") +
  geom_spatvector(data = sample_points) +
  labs(title = sprintf("Randomly samlped points\nMoran's I: %.2f (p-value: %.4f)", 
                       moran_i$estimate[1],moran_i$p.value)) +                   
  theme_cowplot())

# ggsave2("figures/stratified_sample_points_moranI.png",bg = "white")
```


# 2. Explore pre-fire time series data
## a. Spectral indices

### i. Overall time series
The time series for all points looks like this:
```{r}
# Time series for all points
ggplot(df_filtered) +
  geom_point(aes(x = date,y = DailyMeanNDMI),alpha = .2) +
  labs(y = "Daily mean NDMI\n (per point location)") +
  theme_cowplot()

# ggsave2("figures/Timeseries_NDMI_randompoints.png",
#         bg = "white",width = 10, height = 8)
```

### ii. individual point's TS
A look at individual time series reveals diverse trends before the fire:
```{r}
# Time series for some random points
rand_id <- sample(unique(df_filtered$ObservationID), 16)
# rand_id <- c(546,1568,1620,1824,1892,2122,2391,2552,2558,2573,2645,2869,2885,2988,3017,3085)

index_name <- "NDMI"

df_filtered %>% 
  filter(ObservationID %in% rand_id) %>%
  mutate(ObservationID = as.factor(ObservationID)) %>% 
  mutate(ObservationID = fct_reorder(ObservationID, dNBR)) %>% 
  ggplot() +
    geom_point(aes(x = date,y = !!sym(paste0("DailyMean",index_name)),
                   color = BeforeBurnDate),alpha = .2) +
    labs(y = sprintf("Daily mean %s\n (per point location)",index_name),
         title = "Lowest severity top left to highest bottom right") +
    geom_vline(aes(xintercept = as.numeric(burn_date)), color = "red", linetype = "dashed") +
    facet_wrap(~ObservationID) +
    theme_cowplot()

# ggsave2(sprintf("figures/%s_perpoint.png",index_name),
#         bg = "white",width = 9, height = 8)
```

### iii. Histogram of spectral index values
The histogram for this spectral index:
```{r}
mean_index <- df_filtered %>% 
  select(contains(index_name)) %>% 
  pull() %>% mean(.,na.rm = T)

# Histogram of index values
ggplot(df_filtered) +
  geom_histogram(aes(x = !!sym(paste0("DailyMean",index_name)))) +
  geom_vline(xintercept = mean_index,
             color = "red", linetype = "dashed", size = 1) + 
  geom_text(x = 0.2, y = 1e3, 
                label = paste("Mean = ", 
                              round(mean_index, 2)), 
            color = "red", vjust = -1) +
  labs(title = sprintf("Histogram of %s values",index_name)) +
  theme_cowplot()

# ggsave2(sprintf("figures/Histogram_%s.png",index_name),
#         bg = "white",width = 10, height = 8)
```

### iv. Scatter dNBR vs. spectral index

```{r}
# Scatterplot spectral index vs.dNBR 
(ggplot(df_filtered) +
  geom_point(aes(x = !!sym(paste0("DailyMean",index_name)),y = dNBR),alpha = .2) +
  labs(x = sprintf("Daily mean %s\n (per point location)",index_name),
       y = "dNBR") +
 theme_cowplot())
```


## b. LST
The time series of LST for this fire event looks like this:
```{r}
# Plot LST time series
df_filtered %>% 
  filter(ObservationID %in% rand_id) %>%
  mutate(ObservationID = as.factor(ObservationID)) %>% 
  mutate(ObservationID = fct_reorder(ObservationID, dNBR)) %>% 
  ggplot() +
  geom_point(aes(x = date,y = LST, color = BeforeBurnDate),alpha = .2) +
  labs(y = sprintf("Daily mean %s\n (per point location)","LST"),
       title =  "Lowest severity top left to highest bottom right") +
  geom_vline(aes(xintercept = as.numeric(burn_date)), color = "red", linetype = "dashed") +
  facet_wrap(~ObservationID) +
  theme_cowplot()

# ggsave2("figures/LST_perpoint.png",bg = "white",width = 10, height = 8)
```


# 3. Fit curves to pre-fire data

## a. Fit linear regression to pre-fire data
```{r}
# Apply OLS to each time series point
df_mod <- df_filtered %>%
  group_by(ObservationID) %>% 
  filter(BeforeBurnDate) %>% 
  do(mod = lm(!!sym(paste0("DailyMean",index_name)) ~ date, data = .)) %>%
  mutate(Slope = summary(mod)$coeff[2],
         Intercept = summary(mod)$coeff[1]) %>%
  select(-mod)

df_mod <- df_filtered %>%
  group_by(ObservationID) %>%
  select(dNBR) %>% 
  summarize(dNBR = first(dNBR)) %>% 
  select(-ObservationID) %>% 
  cbind(df_mod)
```

Scatterplot of pre-fire NDMI slope vs. dNBR
```{r}
# Plot slope of index vs. dNBR
ggplot(df_mod) +
  geom_point(aes(x = Slope,y = dNBR),alpha = .2) +
  labs(x = sprintf("Slope of pre-fire %s",index_name),
       y = "dNBR") +
  theme_cowplot()

# ggsave2(sprintf("figures/%s_vs_dNBR.png",index_name),
#         bg = "white",width = 10, height = 8)
```

Is there a spatial pattern in the OLS coefficients?
```{r}
sample_points_merged <- sample_points %>% 
  filter(ObservationID %in% df_mod$ObservationID) %>% 
  merge(., df_mod, by = "ObservationID") 

(p <- ggplot() +
    geom_spatraster(data = dnbr_in_perimeter,aes(fill = dNBR)) +
    geom_spatvector(data = sample_points_merged, aes(color = Slope)) +
    scale_fill_viridis_c(option = "magma",
                         na.value = "transparent") +
    scale_color_viridis_c(option = "viridis",
                         na.value = "transparent") +
    labs(fill = "dNBR",
         color = sprintf("Slope %s vs. dNBR",index_name)) +
    theme_cowplot()
)

# ggsave2(p, 
#         filename = sprintf("figures/Map_Slope_per_Point_%s_vs_dNBR_%sth.png",
#                            pct_cutoff*100,index_name),
#         bg = "white",width = 10, height = 8)

```
## b. Splines

Setting up user functions.
```{r}
model_fit_smoothedspline <- function(x,y,spar = 0.5) {
  # Using a spline smoother
  smooth.spline(x = x, y = y, spar = spar)
}

model_index_smoothedspline <- function(x,y,full_data, spar = 0.5) {
  
  # Use function to fit model
  model <- model_fit_smoothedspline(x,y,spar)
  
  # Generate predictions for curve plotting (for time-period doy 130-300) 
  pred <- predict(model, data.frame(doy = 130:300))
  # pred <- unlist(pred$y)
  
  # use function to find vertex (linear model)
  # vertex <- find_vertex(model)
  # find vertex based on predictions (spline smoother)
  vertex <- data.frame(
    x = pred$x[pred$y == max(pred$y)], 
    y = pred$y[pred$y == max(pred$y)]
  )
  
  # Write necessary values back to df
  data <- suppressMessages(full_join(full_data, 
                                     data.frame(
                                       doy = 130:300,
                                       spline.max = vertex$y,
                                       spline.max.doy = vertex$x,
                                       spline.pred = pred))
                           ) 
  return(data)
}
```

We can fit a spline for the day of year and spectral index
```{r}
df_filtered$doy <- lubridate::yday(df_filtered$date)

df <- df_filtered %>% 
  filter(ObservationID == 2391,
         !is.na(DailyMeanNDMI))

test_smooth <- model_index_smoothedspline(df$doy,df$DailyMeanNDMI,df)

df <- df_filtered %>% 
  filter(ObservationID == 2391,
         BeforeBurnDate == TRUE,
         !is.na(DailyMeanNDMI))

test_smooth_prefire <- model_index_smoothedspline(df$doy,df$DailyMeanNDMI,df) %>% 
  rename(spline.pred.prefire = 'spline.pred.doy.1')

test_smooth_join <- test_smooth %>%
  rename(spline.pred = 'spline.pred.doy.1') %>% 
  left_join(test_smooth_prefire %>% select(spline.pred.doy,spline.pred.prefire),
            by = "spline.pred.doy")

ggplot(test_smooth_join) + 
  geom_line(aes(x = doy, y = spline.pred)) +
  geom_line(aes(x = doy, y = spline.pred.prefire),color = "red") +
  geom_point(aes(x = doy, y = DailyMeanNDMI,color = BeforeBurnDate)) +
  labs(y = sprintf("Daily mean %s",index_name),
       subtitle = sprintf("Spline parameter = %s",0.5)) +
  geom_vline(aes(xintercept = yday(burn_date)), color = "red", linetype = "dashed") +
  ylim(c(-0.3,0.3)) +
  theme_cowplot()
```
As we can see, the spline struggles to deal with the outlier at the edge. This outlier from 28 May 2020 is due to the pixel being treated as a valid one (not masked out by Fmask) even though a quick check revealed it was partly snow covered. So maybe an additional snow mask is needed to remove this issue. For now, I'll continue with the outlier in the data.

So maybe a different spline parameter enabling a broader segmentation might help. So I tried spar = 0.7

```{r}
spar <- 0.7

df <- df_filtered %>% 
  filter(ObservationID == 2391,
         !is.na(DailyMeanNDMI))

test_smooth <- model_index_smoothedspline(df$doy,df$DailyMeanNDMI,df,spar = spar)

df <- df_filtered %>% 
  filter(ObservationID == 2391,
         BeforeBurnDate == TRUE,
         !is.na(DailyMeanNDMI))

test_smooth_prefire <- model_index_smoothedspline(df$doy,df$DailyMeanNDMI,df,spar = spar) %>% 
  rename(spline.pred.prefire = 'spline.pred.doy.1')

test_smooth_join <- test_smooth %>%
  rename(spline.pred = 'spline.pred.doy.1') %>% 
  left_join(test_smooth_prefire %>% select(spline.pred.doy,spline.pred.prefire),
            by = "spline.pred.doy")

ggplot(test_smooth_join) + 
  geom_line(aes(x = doy, y = spline.pred)) +
  geom_line(aes(x = doy, y = spline.pred.prefire),color = "red") +
  geom_point(aes(x = doy, y = DailyMeanNDMI,color = BeforeBurnDate)) +
  labs(y = sprintf("Daily mean %s",index_name),
       subtitle = sprintf("Spline parameter = %s",spar)) +
  geom_vline(aes(xintercept = yday(burn_date)), color = "red", linetype = "dashed") +
  ylim(c(-0.3,0.3)) +
  theme_cowplot()
```


```{r}
spar <- 0.7

ndmi_smooth <- df_filtered %>%
  filter(ObservationID %in% rand_id,
         !is.na(DailyMeanNDMI)) %>% 
  group_by(ObservationID) %>% 
  group_modify(~model_index_smoothedspline(.x$doy,.x$DailyMeanNDMI,.x,spar = spar))

ndmi_smooth <- ndmi_smooth %>% 
  rename(spline.pred = 'spline.pred.doy.1')

# Plot multiple time series with spline
(ndmi_smooth %>% 
  ungroup() %>% 
  mutate(ObservationID = as.factor(ObservationID)) %>% 
  mutate(ObservationID = fct_reorder(ObservationID, dNBR)) %>% 
  ggplot() +
  geom_line(aes(x = doy, y = spline.pred)) +
  geom_point(aes(x = doy,y = DailyMeanNDMI, color = BeforeBurnDate),alpha = .2) +
  labs(y = sprintf("Daily mean %s\n (per point location)",index_name),
       title = "Lowest severity top left to highest bottom right",
       subtitle = sprintf("Spline parameter = %s",spar)) +
  geom_vline(aes(xintercept = yday(burn_date)), color = "red", linetype = "dashed") +
  facet_wrap(~ObservationID) +
  theme_cowplot())

# ggsave2(sprintf("figures/%s_perpoint_splines_%s.png",index_name,spar),
#         bg = "white",width = 10, height = 8)
```
It's much better than with the default parameter of 0.5 but still suffers from edge effects and could be too smooth. Some curves look like they're cut short at the time of burning.

So I tried LOESS smoothing, which can deal with edge effects.

## c. LOESS
First, I ran a first order LOESS, so it's fitting a linear regression in each local window:

```{r}
df <- df_filtered %>% filter(ObservationID == 2391)

spans <- c(0.1,0.3,0.4)
lobj1 <- loess.smooth(df$doy,df$DailyMeanNDMI,span = spans[1]) 
lobj2 <- loess.smooth(df$doy,df$DailyMeanNDMI,span = spans[2])
lobj3 <- loess.smooth(df$doy,df$DailyMeanNDMI,span = spans[3])

ggplot() +
  geom_point(data = df,aes(x = doy, y = DailyMeanNDMI)) + 
  geom_line(aes(x = lobj1$x,y = lobj1$y,
                color = sprintf("span = %s",spans[1]))) +
  geom_line(aes(x = lobj2$x,y = lobj2$y,
                color = sprintf("span = %s",spans[2]))) +
  geom_line(aes(x = lobj3$x,y = lobj3$y, 
                color = sprintf("span = %s",spans[3]))) +
  geom_vline(aes(xintercept = yday(df$burn_date)), color = "red", linetype = "dashed") +
  labs(y = sprintf("Daily mean %s\n (per point location)",index_name),
       color = "Smoothness parameters",
       title = "LOESS fits",
       subtitle = "(local polynomial, p = 1)") +
  theme_cowplot()

# ggsave2(sprintf("figures/%s_LOESS.png",index_name),
#         bg = "white",width = 10, height = 8)
```
Would a second order polynomial be better?
```{r}
df <- df_filtered %>% filter(ObservationID == 2391)

spans <- c(0.1,0.3,0.4)
lobj1 <- loess.smooth(df$doy,df$DailyMeanNDMI,span = spans[1], degree = 2) 
lobj2 <- loess.smooth(df$doy,df$DailyMeanNDMI,span = spans[2], degree = 2)
lobj3 <- loess.smooth(df$doy,df$DailyMeanNDMI,span = spans[3], degree = 2)

ggplot() +
  geom_point(data = df,aes(x = doy, y = DailyMeanNDMI)) + 
  geom_line(aes(x = lobj1$x,y = lobj1$y,
                color = sprintf("span = %s",spans[1]))) +
  geom_line(aes(x = lobj2$x,y = lobj2$y,
                color = sprintf("span = %s",spans[2]))) +
  geom_line(aes(x = lobj3$x,y = lobj3$y, 
                color = sprintf("span = %s",spans[3]))) +
  geom_vline(aes(xintercept = yday(df$burn_date)), color = "red", linetype = "dashed") +
  labs(y = sprintf("Daily mean %s\n (per point location)",index_name),
       color = "Smoothness parameters",
       title = "LOESS fits",
       subtitle = "(local polynomial, p = 2)") +
  theme_cowplot()

```

# 4. Retrieve time-integrated metric
I use the full-season splines (spar = 0.5) to predict NDMI values at different time intervals before the fire and then calculate their integrals.

```{r}
df_filtered$doy <- lubridate::yday(df_filtered$date)

df <- df_filtered %>% 
  filter(ObservationID == 2391,
         !is.na(DailyMeanNDMI))

test_smooth <- model_index_smoothedspline(df$doy,df$DailyMeanNDMI,df) %>% 
  rename(spline.pred = 'spline.pred.doy.1')

model <- model_fit_smoothedspline(df$doy,df$DailyMeanNDMI,spar=0.5)
pred <- data.frame(predict(model, data.frame(doy = 130:300))) %>% 
  rename(doy = doy, index = doy.1)

# Get DOY of burn
doy_burn <- yday(df$burn_date[1])

pred_before_burn <- pred %>% 
  filter(doy < doy_burn) %>% 
  mutate(TI_index = revcumsum(index),
         days_before_fire = doy_burn - doy)

TI_1week <- pred_before_burn %>% 
  filter(days_before_fire ==7)

```


```{r}
# Restructured function to return vector of predictions
model_index_smoothedspline <- function(x,y,full_data, spar = 0.5) {
  
  # Use function to fit model
  model <- model_fit_smoothedspline(x,y,spar)
  
  # Generate predictions for curve plotting (for time-period doy 130-300) 
  pred <- data.frame(predict(model, data.frame(doy = 130:300))) %>% 
  rename(doy = doy, index = doy.1)

  # Get DOY of burn
  doy_burn <- yday(full_data$burn_date[1])
  
  pred_before_burn <- pred %>% 
    filter(doy < doy_burn) %>% 
    mutate(TI_index = revcumsum(index),
           days_before_fire = doy_burn - doy)
  
  # Prepare output vector of 40-day predictions
  out_data <- pred_before_burn %>% 
    filter(days_before_fire <= 40) %>% 
    select(days_before_fire, TI_index) %>% 
    pivot_wider(
      names_from = days_before_fire,
      values_from = TI_index,
      names_prefix = "d_prefire_"
      ) 
    
  return(out_data)
}

spar <- 0.5

# Apply spline to each time series point
ndmi_smooth <- df_filtered %>%
  filter(!is.na(DailyMeanNDMI)) %>% 
  group_by(ObservationID) %>% 
  group_modify(~model_index_smoothedspline(.x$doy,.x$DailyMeanNDMI,.x,spar = spar))


# Merge dataframes
data <- suppressMessages(df_filtered %>%
                           # filter(ObservationID %in% rand_id) %>% 
                           select(ObservationID,dNBR,burn_date) %>% 
                           group_by(ObservationID) %>% 
                           unique() %>% 
                           ungroup() %>% 
                           right_join(ndmi_smooth, by = "ObservationID")
                         ) 

```

```{r}
# Plot time-integrated NDMI 7 days before fire vs. dNBR
ggplot(data) +
  geom_point(aes(x = d_prefire_7 ,y = dNBR),alpha = .2) +
  labs(x = sprintf("time-integrated NDMI 7 days pre-fire %s",index_name),
       y = "dNBR") +
  theme_cowplot()
```
Now, let's see which time before the fire has the highest R2

```{r}
d_prefire_cols <- grep("^d_prefire_", names(data), value = TRUE)

# Function to run lm and extract statistics
run_lm <- function(y_var, x_var, data) {
  formula <- as.formula(paste(y_var, "~", x_var))
  model <- lm(formula, data = data)
  summary <- summary(model)
  
  # get days before fire as numeric
  days_before <- as.numeric(str_extract(x_var, "\\d+"))
  
  tibble(
    predictor = x_var,
    days_before_fire = days_before,
    r_squared = summary$r.squared,
    adj_r_squared = summary$adj.r.squared,
    coefficient = coef(model)[2],
    p_value = summary$coefficients[2, 4]
  )
}

# Run lm for each d_prefire column and combine results
results <- map_dfr(d_prefire_cols, ~run_lm("dNBR", .x, data))

results <- results %>% arrange(days_before_fire)

ggplot(results) + 
  geom_point(aes(x = days_before_fire, y = r_squared)) + 
  theme_cowplot()

mean_ndmi_pre_day <- colMeans(data[d_prefire_cols])
results %>% 
  mutate(avg_ndmi = mean_ndmi_pre_day) %>% 
  ggplot() + 
  geom_point(aes(x = avg_ndmi, y = r_squared, color = days_before_fire)) + 
  scale_color_viridis_c(option = "inferno") +
  theme_cowplot()
```

